sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Table 1 links

Systems for specifying statistical models that can include the SPDE, such as \pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as \pkg{lme4} [@bates2015] and \pkg{glmmTMB} [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
We provide a more detailed comparison of related software packages in Table 1 and the Discussion.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table 1).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, \pkg{VAST} is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table 1).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).

@@

\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.}
\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item $^1$Technically possible but non-trivial. $^2$Penalized smoother GAMs that determine `wiggliness'. $^3$\pkg{inlabru} but not \pkg{INLA}. $^4$Spatiotemporal fields as random walk only. $^5$SPDE approach as in Miller et al. (2019). $^6$Does have predictive process knots. $^7$Zero-inflated NB2 only. $^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}. $^9$Possible as log-linked Poisson GLMM with aggregated data. $^{10}$Hurdle models possible by fitting components separately.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

@@

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Penalized spline support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.

>>>>

Systems for specifying statistical models that can include the SPDE, such as \pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as \pkg{lme4} [@bates2015] and \pkg{glmmTMB} [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
<!-- However, \pkg{VAST} is typically less effective at replicating nonlinear preferences than GAMs, which are supported by several of these packages [@brodie2020]. -->
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} or \pkg{inlabru} (Appendix 4) and \pkg{VAST} (Appendix 5).

@@

\caption{Comparison of functionality between several \proglang{R} packages that can fit geostatistical GLMMs.}
\centering
\fontsize{8}{10}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lllllll}
\label{tab:functionality}
\toprule
 & \pkg{sdmTMB} & \pkg{VAST} & \pkg{INLA}/\pkg{inlabru} & \pkg{mgcv} & \pkg{spBayes} & \pkg{spaMM}\\
\midrule
Time-varying coefficients & \checkmark & --$^1$ & \checkmark & \checkmark & \checkmark & --\\
Spatially varying coefficients (SVC) & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
GAMs$^2$ & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Threshold covariates & \checkmark & -- & \checkmark$^3$ & -- & -- & --\\
Offsets & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark\\
\addlinespace
Spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark$^4$ & --\\
Spatial + spatiotemporal fields & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Anisotropy & \checkmark & \checkmark & -- & \checkmark & -- & --\\
Correlation barriers & \checkmark & \checkmark & \checkmark & \checkmark & -- & --\\
Separate range parameters for fields & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Share range parameters across fields & \checkmark & \checkmark & \checkmark & -- & -- & --\\
SPDE-based & \checkmark & \checkmark & \checkmark & \checkmark$^5$ & --$^6$ & \checkmark\\
\addlinespace
NB1 distribution & \checkmark & -- & \checkmark & \checkmark & -- & \checkmark\\
NB2 distribution & \checkmark & \checkmark$^7$ & \checkmark & \checkmark & -- & \checkmark\\
Zero-truncated distributions & \checkmark & -- & \checkmark & -- & -- & \checkmark\\
Zero-inflated distributions & \checkmark & \checkmark & \checkmark & -- & -- & \checkmark\\
Tweedie distribution & \checkmark & \checkmark & \checkmark & \checkmark$^8$ & -- & --\\
Student-t distribution & \checkmark & -- & \checkmark & \checkmark & -- & --\\
Censored Poisson distribution & \checkmark & -- & \checkmark & -- & -- & --\\
log Gaussian Cox processes & --$^9$ & --$^9$ & \checkmark & --$^9$ & --$^9$ & --$^9$\\
Multivariate responses & -- & \checkmark & \checkmark & -- & \checkmark & \checkmark\\
Built-in delta/hurdle models & \checkmark & \checkmark & \checkmark & --$^{10}$ & -- & \checkmark\\
Poisson-link delta model & \checkmark & \checkmark & \checkmark & -- & -- & --\\
Likelihood weights & \checkmark & -- & \checkmark & \checkmark & \checkmark & \checkmark\\
Maximum/marginal likelihood & \checkmark & \checkmark & -- & \checkmark & -- & --\\
\addlinespace
Bayesian/optionally Bayesian & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & --\\
Priors/penalties & \checkmark & -- & \checkmark & -- & \checkmark & --\\
Matern PC priors & \checkmark & -- & \checkmark & -- & -- & --\\
\addlinespace
Spatial (or spatial dynamic) factor analysis & -- & \checkmark & -- & -- & -- & --\\
Empirical Orthogonal Function (EOF) analysis & -- & \checkmark & -- & -- & -- & --\\
Built-in area-weighted index standardization & \checkmark & \checkmark & -- & -- & -- & --\\
Built-in cross-validation & \checkmark & -- & -- & -- & -- & --\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item $^1$Technically possible but non-trivial. $^2$Penalized smoother GAMs that determine `wiggliness'. $^3$\pkg{inlabru} but not \pkg{INLA}. $^4$Spatiotemporal fields as random walk only. $^5$SPDE approach as in Miller et al. (2019). $^6$Does have predictive process knots. $^7$Zero-inflated NB2 only. $^8$Tweedie power parameter fixed for \texttt{mgcv::gamm()}. $^9$Possible as log-linked Poisson GLMM with aggregated data. $^{10}$Hurdle models possible by fitting components separately.
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

@@

\pkg{sdmTMB} would not be possible without the \pkg{TMB} [@kristensen2016] and \pkg{INLA} [@rue2009; @lindgren2011; @lindgren2015] \proglang{R} packages.
\pkg{sdmTMB} is heavily inspired by and in some places code has been adapted from both the \pkg{VAST} [@thorson2019] and \pkg{glmmTMB} [@brooks2017] \proglang{R} packages.
Smoother support was possible thanks to \pkg{mgcv} [@wood2017a].
We thank the authors of all these packages.
We thank S. Kotwicki, M. Lindmark, C.C. Monnahan, P.M. Regular, J.T. Thorson, and J. Watson for helpful comments that substantially improved the manuscript.
